The Key to a Data Parallel Compiler

### 2016-04-05 2016-spring-array/2016-array-key.tex
Add in most of the other sections, tables, and other goodies.

\maketitle
\begin{abstract}
TBW
\end{abstract}

%[c]Proposed new outline:
%[c]
%[c]I. 	Generalization on manipulations over Sub-tree computation
%[c]  A.	Key + Encoding → Data parallel, Array-based computation
%[c]  B.	Easy to write/read
%[c]  C.	Uniform across passes
%[c]II.	Case Studies
%[c]  A.	Function Lifting
%[c]  B.	Expression Flattening
%[c]  C.	Lexical Scope Resolution
%[c]  D.	Illustrates the problem space
%[c]III.	Other passes/Issues
%[c]  A.	Discussion of other passes
%[c]  B.	Beyond the Key Operator
%[c]IV.	In practice/Co-dfns Compiler
%[c]V.	Performance/Results
%[c]
%[c]Questions that must be addressed:
%[c]
%[c]1. Why the formalization in APL?
%[c]2. Why parallelize a compiler?
%[c]3. How well does it work/scale/generalize

%[of]:Introduction
\section{Introduction}

@@

\begin{itemize}[noitemsep]
\item We describe a method of computing over arbitrary sub-trees selected by their parent-child 
relationships in a data-parallel manner using the Key operator and Node Coordinates.
\item We situate this technique into a broader, language-driven compiler construction strategy 
that enables the development of parallel by construction compilers that exhibit desirable programmability 
features in terms of code size, complexity, portability, and maintainability. 
\item We provide analysis of these techniques as used in a commercial compiler project called Co-dfns and 
report on the results.
\item We compare these techniques to the Nanopass approach to compiler design in terms of programmer 
productivity and compiler runtime performance. 
\item We enumerate a set of programming techniques, patterns, and strategies that complement the 
Key operator and Node Coordinates in the construction of tree transformations in general and give some 
intuitions on the use of these techniques on arbitrary transformations.
\end{itemize}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol                   & Monadic            & Dyadic \\
\midrule
\texttt{+}               & Identity           & Plus (Add) \\
\texttt{-}               & Negative           & Minus (Subtract) \\
\texttt{×}               & Direction (Signum) & Times (Multiply) \\
\texttt{÷}               & Reciprocal         & Divide \\
\texttt{|}               & Magnitude          & Residue (Modulo) \\
\texttt{⌊}               & Floor              & Minimum \\
\texttt{⌈}               & Ceiling            & Maximum \\
\texttt{*}               & Exponential        & Power \\
\texttt{⍟}               & Natural Logarithm  & Logarithm \\
\texttt{○}               & Pi Times           & Circular (Trigonometric) \\
\texttt{!}               & Factorial          & Binomial \\
\texttt{\textasciitilde} & Not                & \\
\texttt{?}               & Roll               & \\
\texttt{\^}              &                    & And \\
\texttt{∨}               &                    & Or \\
\texttt{⍲}               &                    & Nand \\
\texttt{⍱}               &                    & Nor \\
\texttt{<}               &                    & Less \\
\texttt{≤}               &                    & Less Or Equal \\
\texttt{=}               &                    & Equal \\
\texttt{≥}               &                    & Greater Or Equal \\
\texttt{>}               &                    & Greater \\
\texttt{≠}               &                    & Not Equal \\
\end{tabular}
\caption{Scalar Primitives in Proposed Language}
\label{tab:scalarprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic       & Dyadic \\
\midrule
\texttt{⍴} & Reshape       & \\
\texttt{,} & Ravel         & Catenate/Laminate \\
\texttt{⍪} & Table         & Catenate First/Laminate \\
\texttt{⌽} & Reverse       & Rotate \\
\texttt{⊖} & Reverse First & Rotate First \\
\texttt{⍉} & Transpose     & Transpose \\
\texttt{↑} & Mix           & \\
\texttt{↓} & Split         & \\
\texttt{⊂} & Enclose       & Partitioned Enclose \\
\texttt{∊} & Enlist        & \\
\end{tabular}
\caption{Structural Mixed Primitives in Proposed Language}
\label{tab:structprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol                   & Monadic  & Dyadic \\
\midrule
\texttt{⊃}               & Disclose & Pick\\
\texttt{↑}               &          & Take \\
\texttt{↓}               &          & Drop \\
\texttt{/}               &          & Replicate \\
\texttt{⌿}               &          & Replicate First \\
\texttt{\textbackslash}  &          & Expand \\
\texttt{⍀}               &          & Expand First \\
\texttt{\textasciitilde} &          & Without (Excluding) \\
\texttt{∩}               &          & Intersection \\
\texttt{∪}               & Unique   & Union \\
\texttt{⊣}               & Same     & Left \\
\texttt{⊢}               & Identity & Right\\
\end{tabular}
\caption{Selection Mixed Primitives in Proposed Language}
\label{tab:selprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic         & Dyadic \\
\midrule
\texttt{⍳} & Index Generator & Index Of \\
\texttt{∊} &                 & Membership \\
\texttt{⍋} & Grade Up        & Grade Up \\
\texttt{⍒} & Grade Down      & Grade Down \\
\texttt{?} &                 & Deal \\
\texttt{⍷} &                 & Find \\
\end{tabular}
\caption{Selector Mixed Primitives in Proposed Language}
\label{tab:seltorprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic       & Dyadic \\
\midrule
\texttt{⍴} & Shape         & \\
\texttt{≡} & Depth         & Match \\
\texttt{≢} & Tally         & Not Match \\
\texttt{⍎} & Execute       & Execute \\
\texttt{⍕} & Format        & Format \\
\texttt{⊥} &               & Decode (Base) \\
\texttt{⊤} &               & Encode (Representation) \\
\texttt{⌹} & Matrix Divide & Matrix Inverse \\
\end{tabular}
\caption{Miscellaneous Mixed Primitives in Proposed Language}
\label{tab:miscprims}
\end{table}

\begin{table*}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Name & Description \\
\midrule
\texttt{⍨} & Commute & Swaps arguments or distributes
 right argument to both sides \\
\texttt{¨} & Each & Applies its operand point-wise over the left/right
 arguments \\
\texttt{/} & Reduce & Reduce along the last axis \\
\texttt{⌿} & Reduce First & Reduce along the first axis \\
\texttt{\textbackslash} & Scan & Scan along the last axis \\
\texttt{⍀} & Scan First & Scan along the first axis \\
\texttt{⌸} & Key & Apply operand once for each sub-array grouped by key \\
\end{tabular}
\caption{Primitive Adverbs in Proposed Language, each takes a single left
operand and describes a verb operating over one or two arguments}
\label{tab:adverbs}
\end{table*}

\begin{table*}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Name & Description \\
\midrule
\texttt{∘} & Compose & Composes two operands as in traditional mathematics \\
\texttt{.} & Inner Product & Inner product operation, e.g.
 \texttt{+.×} for matrix multiplication \\
\texttt{∘.} & Outer Product & Cartesian product or ``function table'' \\
\texttt{⍣} & Power & Iteration, Limited use only \\
\texttt{⍤} & Rank & Apply a verb along cells of an array \\
\end{tabular}
\caption{Primitive Conjunctions in Proposed Language, each takes a left
 and right operand and describes a verb operating over one or two arguments}
\label{tab:conjunctions}
\end{table*}

@@

%[of]:The Key Operator
\subsection{The Key Operator}

@@

In our case, we use either Fk or Ek as our keys applied to the corresponding AST. We also will drop off 
the first row in each AST using 1↓ since this node “contains” everything. In a complete AST this is 
usually the Module boundary node which contains the entire set of functions and values in the module.
%[cf]
%[of]:Encoding the AST
\subsection{Encoding the AST}

@@

Note especially that all inter-node information lives in the depth vector, but that this information requires 
non-local access as described in the previous section to use it. Constructing node coordinates fixes this issue.
%[cf]
%[of]:Node Coordinates
\subsection{Node Coordinates}

@@

%[of]:Constructing Node Coordinates
\subsubsection{Constructing Node Coordinates}

@@

A careful study of the definition of a node coordinate and the above construction should 
reveal why this works. Intuitively, we are creating a multi-dimensional space or a number 
system in which each digit place or dimension contains or circumscribes a smaller space in 
which are contained all the descendant nodes that appear lower in the tree. Each coordinate 
is a sort of special path through the tree encoded to have desirable properties relative to 
other paths.
%[cf]
%[of]:Operations on Node Coordinates
\subsubsection{Operations on Node Coordinates}

@@

\begin{verbatim}
A(f g h)B ←→ (A f B) g (A h B)
A(0 f h)B ←→ 0 f (A h B) ⍝ Any constant for 0
A(f g)B   ←→ f (A g B)
\end{verbatim}

@@

We will use these keys to perform computation over the AST and accomplish our tasks of function lifting and 
expression flattening in the next section, but we make a final note here, that we can imagine many other 
operations which might be used throughout the compiler when we care about how nodes relate to each 
other. The use of operators like reduction, scan, inner and outer products allow us to obtain the necessary 
information from the node coordinate matrix. Sometimes these results might not be obvious, but the 
example of the “belongs to” relationship given above demonstrates a pattern that arises repeatedly 
throughout our data parallel compiler, and is therefore particularly useful to understand. 
%[cf]

%[cf]
%[cf]
%[of]:Case Studies
\section{Case Studies}

%[of]:Function Lifting
\subsection{Function Lifting}

@@

The final task to complete function lifting of each function body before lifting is to shift the depths in 
the depth vectors to correspond to those of a function lifted to the top-level, and to attach a function 
node to the top of each of the bodies. At that point, we simply recombine all of the newly created functions 
into a single top level. 
%[cf]
%[of]:Expression Flattening
\subsection{Expression Flattening}

@@

The only extra issue involved here is to ensure that the order of evaluation matches. In our case, we 
are assuming that the order of evaluation is right to left, which means that the above order is actually 
backwards of the desired order. During recombination, we simply reverse these orders and this fixes 
that problem. More work would be required to take into consideration a specific precedence hierarchy. 
%[cf]
%[of]:Lexical Scoping
\subsection{Lexical Scoping}
%[cf]
%[cf]
%[of]:Table of Compiler Passes
\renewcommand{\arraystretch}{1.5}
\begin{table*}
\centering
\begin{tabular}{l p{2.8in} l l l}
\toprule
Pass & Description & Core Patterns & Coordinate? \\
\midrule
Record Node Coordinates & Adds a new field to each node containing that node's 
 node coordinate. & Outer Product/Scan & Yes \\
Record Function Depths & Adds a new field to each node recording how many 
 functions surround the node. & Inner Product & Yes \\
Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
 the top-level. & Filter & No \\
Drop Unreachable Code & Eliminates some unreachable code. &
 Filter/Inner Product & Yes \\
Lift Functions & Moves all functions to the top-level. & Key & Yes \\
Drop Redundant Nodes & Eliminates unnecessary nodes/nesting & Filter & No \\
Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
Compress Atomic Nodes & Atomizes nested atom nodes & Amend & No \\
Propagate Constants & Inlines all references to literal values. &
 Amend/Rank & Yes \\
Fold Constants & Converts constant expressions to literals & Amend & No \\
Compress Expressions & Converts expression sub-trees into single nodes. & 
 Amend & No \\
Record Final Return Value & Records the value returned by each function. & 
 Key & No \\
Normalize Values Field & Normalizes the shape and size of the values field. &
 Amend & No \\
Lift Type-checking & Infers some type information at compile time &
 Power Limit/Rank & No \\
Allocate Value Slots & Does a form of frame allocation for variables &
 Key & No \\
Anchor Variables & Resolves lexically scoped variables &
 Key & Yes \\
Record Live Variables & Records the variables that are live at each point of 
 execution. & Key & No \\
Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions &
 Key & No \\
Type Specialization & Specializes each function for a series of potential inputs. 
 & Key & No \\
\end{tabular}
\caption{A listing of some compiler passes in the Co-dfns compiler and their 
 relationship with the Key operator and associated tree computation techniques}
\label{table:passes}
\end{table*}
%[cf]
%[of]:Other Passes
\section{Other Passes}

Table \ref{table:passes} contains a selection of passes from the Co-dfns compiler 
and breaks down their features based on how they traverse the tree and whether 
or not they use the Key operator or Node Coordinates. They are oriented with 
front-end passes higher in the table and passes further down the compiler 
chain at the bottom of the table. These passes give a good idea of how the Key 
operator plays a part in the complete compiler as well as how Node Coordinates 
help to deal with complex sub-tree selections.

@@

In short, the goal of the structure and design of the compiler passes given 
in table \ref{table:passes} is to utilize Node Coordinates to remove as quickly as possible 
the nested structure of the AST into a flatter form that does not require complex 
analysis to traverse. 
%[cf]
%[of]:The Co-dfns Compiler
\section{The Co-dfns Compiler}
%[cf]
%[of]:Performance
\section{Performance}
%[cf]
%[of]:Future Work
\section{Future Work}
%[cf]
%[of]:Related Work
\section{Related Work}

@@

J. D. Bunda and J. A. Gerth presented a method for doing table driven parsing of APL which 
suggested a parallel optimization for parsing, but did not elucidate the algorithm [7].
%[cf]
%[of]:Conclusion
\section{Conclusion}

@@

We have demonstrated the technique and the core insights behind the data structures involved. It 
presents a solution to a very old and traditional problem in a very uncommon light, by eschewing 
the common practices that underlie every other significant and general solution found in modern 
compilers today and replacing them with an entirely different paradigm centred on parallelism 
and aggregate operations.
%[cf]
%[of]:Acknowledgements
%[cf]

>>>>

\maketitle
\begin{abstract}
We present a language-driven strategy for the construction of 
compilers that are inherently data-parallel in their design and 
implementation. Using an encoding of the inter-node relationships 
between nodes in an AST called a Node Coordinate Matrix, we demonstrate how 
an operator called the Key operator, that applies a function over groupings 
of array cells grouped and ordered by their keys, when used in conjunction 
with a Node Coordinate Matrix, permits arbitrary computation over sub-trees 
of an AST using purely data-parallel array programming techniques. We discuss 
the use of these techniques when applied to an experimental commercial 
compiler called Co-dfns, which is a fully data-parallel compiler developed 
using the techniques discussed here.
\end{abstract}

% Proposed new outline:
% 
% I. 	Generalization on manipulations over Sub-tree computation
%   A.	Key + Encoding → Data parallel, Array-based computation
%   B.	Easy to write/read
%   C.	Uniform across passes
% II.	Case Studies
%   A.	Function Lifting
%   B.	Expression Flattening
%   C.	Lexical Scope Resolution
%   D.	Illustrates the problem space
% III.	Other passes/Issues
%   A.	Discussion of other passes
%   B.	Beyond the Key Operator
% IV.	In practice/Co-dfns Compiler
% V.	Performance/Results
% 
% Questions that must be addressed:
% 
% 1. Why the formalization in APL?
% 2. Why parallelize a compiler?
% 3. How well does it work/scale/generalize

\section{Introduction}

@@

\begin{itemize}[noitemsep]

\item We describe a method of computing over arbitrary sub-trees
selected by their parent-child relationships in a data-parallel manner
using the Key operator and Node Coordinates.

\item We situate this technique into a broader, language-driven
compiler construction strategy that enables the development of parallel
by construction compilers that exhibit desirable programmability
features in terms of code size, complexity, portability, and
maintainability.

\item We provide analysis of these techniques as used in a commercial
compiler project called Co-dfns and report on the results.

\item We compare these techniques to the Nanopass approach to
compiler design in terms of programmer productivity and compiler
runtime performance.

\item We enumerate a set of programming techniques, patterns, and
strategies that complement the Key operator and Node Coordinates
in the construction of tree transformations in general and give some
intuitions on the use of these techniques on arbitrary transformations.
\end{itemize}

\section{Notational Conventions}

For space and convenience, we use a concise array notational convention to 
describe our techniques and approach. The notation is executable and well 
established in the array community (it is a limited subset of APL). This is 
in fact the same notation used within the Co-dfns compiler itself, whose 
source code is available online and provides a complete example of the use 
of both these algorithms and this notation in the large.

All code examples are given in the following form:

\begin{verbatim}
      5×3+4 ⍝ Right to Left Evaluation
35
\end{verbatim}

That is, the sample code given indented by 6 spaces, followed by the value 
of the expression without indent. All expressions are evaluated right to 
left and all function application is unified and infix; that is, a function 
application may apply a function (such as \verb;+;) to one or two arguments, 
which must appear on the right and (optionally) on the left of the function 
name, as in the above example. Tables \ref{tab:scalarprims} to 
\ref{tab:conjunctions}
provide a listing of all of the primitives used in the Co-dfns compiler, 
a selection of which are used here.

We separate out higher-order functions called operators, from functions that 
operate over arrays that we call simply functions. Operators take a function 
argument or two, and bind more strongly to the left than the right. An 
operator may take one or two operands, either on the left, or on the 
right and left, and will return a function. In the following 
example, we apply a reduction operator to a function created using the 
composition operator, as an example:

\begin{verbatim}
      +∘-/1 2 3 4 5
3
      +/1 2 3 4 5
15
\end{verbatim}

Uses of the notation are introduced slowly throughout the treatment, so 
the reader is encouraged to simply proceed through the treatment, referencing 
the tables only when required.

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol                   & Monadic            & Dyadic \\
\midrule
\texttt{+}               & Identity           & Plus (Add) \\
\texttt{-}               & Negative           & Minus (Subtract) \\
\texttt{×}               & Direction (Signum) & Times (Multiply) \\
\texttt{÷}               & Reciprocal         & Divide \\
\texttt{|}               & Magnitude          & Residue (Modulo) \\
\texttt{⌊}               & Floor              & Minimum \\
\texttt{⌈}               & Ceiling            & Maximum \\
\texttt{*}               & Exponential        & Power \\
\texttt{⍟}               & Natural Logarithm  & Logarithm \\
\texttt{○}               & Pi Times           & Circular (Trigonometric) \\
\texttt{!}               & Factorial          & Binomial \\
\texttt{\textasciitilde} & Not                & \\
\texttt{?}               & Roll               & \\
\texttt{\^}              &                    & And \\
\texttt{∨}               &                    & Or \\
\texttt{⍲}               &                    & Nand \\
\texttt{⍱}               &                    & Nor \\
\texttt{<}               &                    & Less \\
\texttt{≤}               &                    & Less Or Equal \\
\texttt{=}               &                    & Equal \\
\texttt{≥}               &                    & Greater Or Equal \\
\texttt{>}               &                    & Greater \\
\texttt{≠}               &                    & Not Equal \\
\end{tabular}
\caption{Scalar Primitives}
\label{tab:scalarprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic       & Dyadic \\
\midrule
\texttt{⍴} & Reshape       & \\
\texttt{,} & Ravel         & Catenate/Laminate \\
\texttt{⍪} & Table         & Catenate First/Laminate \\
\texttt{⌽} & Reverse       & Rotate \\
\texttt{⊖} & Reverse First & Rotate First \\
\texttt{⍉} & Transpose     & Transpose \\
\texttt{↑} & Mix           & \\
\texttt{↓} & Split         & \\
\texttt{⊂} & Enclose       & Partitioned Enclose \\
\texttt{∊} & Enlist        & \\
\end{tabular}
\caption{Structural Mixed Primitives}
\label{tab:structprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol                   & Monadic  & Dyadic \\
\midrule
\texttt{⊃}               & Disclose & Pick\\
\texttt{↑}               &          & Take \\
\texttt{↓}               &          & Drop \\
\texttt{/}               &          & Replicate \\
\texttt{⌿}               &          & Replicate First \\
\texttt{\textbackslash}  &          & Expand \\
\texttt{⍀}               &          & Expand First \\
\texttt{\textasciitilde} &          & Without (Excluding) \\
\texttt{∩}               &          & Intersection \\
\texttt{∪}               & Unique   & Union \\
\texttt{⊣}               & Same     & Left \\
\texttt{⊢}               & Identity & Right\\
\end{tabular}
\caption{Selection Mixed Primitives}
\label{tab:selprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic         & Dyadic \\
\midrule
\texttt{⍳} & Index Generator & Index Of \\
\texttt{∊} &                 & Membership \\
\texttt{⍋} & Grade Up        & Grade Up \\
\texttt{⍒} & Grade Down      & Grade Down \\
\texttt{?} &                 & Deal \\
\texttt{⍷} &                 & Find \\
\end{tabular}
\caption{Selector Mixed Primitives}
\label{tab:seltorprims}
\end{table}

\begin{table}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Monadic       & Dyadic \\
\midrule
\texttt{⍴} & Shape         & \\
\texttt{≡} & Depth         & Match \\
\texttt{≢} & Tally         & Not Match \\
\texttt{⍎} & Execute       & Execute \\
\texttt{⍕} & Format        & Format \\
\texttt{⊥} &               & Decode (Base) \\
\texttt{⊤} &               & Encode (Representation) \\
\texttt{⌹} & Matrix Divide & Matrix Inverse \\
\end{tabular}
\caption{Miscellaneous Mixed Primitives}
\label{tab:miscprims}
\end{table}

\begin{table*}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Name & Description \\
\midrule
\texttt{⍨} & Commute & Swaps arguments or distributes
 right argument to both sides \\
\texttt{¨} & Each & Applies its operand point-wise over the left/right
 arguments \\
\texttt{/} & Reduce & Reduce along the last axis \\
\texttt{⌿} & Reduce First & Reduce along the first axis \\
\texttt{\textbackslash} & Scan & Scan along the last axis \\
\texttt{⍀} & Scan First & Scan along the first axis \\
\texttt{⌸} & Key & Apply operand once for each sub-array grouped by key \\
\end{tabular}
\caption{Primitive Monadic/Unary Operators, each takes a single left
operand and describes a verb operating over one or two arguments}
\label{tab:adverbs}
\end{table*}

\begin{table*}
\centering
\begin{tabular}{cll}
\toprule
Symbol     & Name & Description \\
\midrule
\texttt{∘} & Compose & Composes two operands as in traditional mathematics \\
\texttt{.} & Inner Product & Inner product operation, e.g.
 \texttt{+.×} for matrix multiplication \\
\texttt{∘.} & Outer Product & Cartesian product or ``function table'' \\
\texttt{⍣} & Power & Iteration, Limited use only \\
\texttt{⍤} & Rank & Apply a verb along cells of an array \\
\end{tabular}
\caption{Primitive Dyadic/Binary Operators, each takes a left
 and right operand and describes a verb operating over one or two arguments}
\label{tab:conjunctions}
\end{table*}

@@

Computing over sub-trees normally involves recursing down the structure of 
the AST, identifying the root nodes of each sub-tree, and then dispatching
to a handler for that sub-tree, which will continue the recursion at that 
point and return the modified tree, which will replace the previous 
sub-tree. More complicated transformations may involved moving the root 
nodes around in the tree or other large, distant structural modifications.

We divide the work of sub-tree computation into three basic phases. Firstly, 
we maintain a node coordinate with each node that uniquely identifies it 
relative to others, we use these coordinates to select and group nodes for 
work as a single sub-tree, and then we operate over these groupings using 
the key operator.

\subsection{The Key Operator}

@@

In our case, we use either Fk or Ek as our keys applied to the
corresponding AST. We also will drop off the first row in each AST
using 1↓ since this node “contains” everything. In a complete
AST this is usually the Module boundary node which contains the entire
set of functions and values in the module.

\subsection{Encoding the AST}

@@

Note especially that all inter-node information lives in the depth
vector, but that this information requires non-local access as
described in the previous section to use it. Constructing node
coordinates fixes this issue.

\subsection{Node Coordinates}

@@

\subsubsection{Constructing Node Coordinates}

@@

A careful study of the definition of a node coordinate and the
above construction should reveal why this works. Intuitively, we are
creating a multi-dimensional space or a number system in which each
digit place or dimension contains or circumscribes a smaller space
in which are contained all the descendant nodes that appear lower in
the tree. Each coordinate is a sort of special path through the tree
encoded to have desirable properties relative to other paths.

\subsubsection{Operations on Node Coordinates}

@@

\begin{verbatim}
A(f g h)B ←→ (A f B) g (A h B)
A(0 f h)B ←→ 0 f (A h B) ⍝ 0 is a constant
A(f g)B   ←→ f (A g B)
\end{verbatim}

@@

We will use these keys to perform computation over the AST and
accomplish our tasks of function lifting and expression flattening in
the next section, but we make a final note here, that we can imagine
many other operations which might be used throughout the compiler when
we care about how nodes relate to each other. The use of operators
like reduction, scan, inner and outer products allow us to obtain
the necessary information from the node coordinate matrix. Sometimes
these results might not be obvious, but the example of the “belongs
to” relationship given above demonstrates a pattern that arises
repeatedly throughout our data parallel compiler, and is therefore
particularly useful to understand.

\begin{table*}
\centering
\begin{tabular}{l p{2.8in} l l l}
\toprule
Pass & Description & Core Patterns & Coordinate? \\
\midrule
Record Node Coordinates & Adds a new field to each node containing that node's 
 node coordinate. & Outer Product/Scan & Yes \\
Record Function Depths & Adds a new field to each node recording how many 
 functions surround the node. & Inner Product & Yes \\
Drop Unnamed Functions & Eliminates some code that will not be evaluated at 
 the top-level. & Filter & No \\
Drop Unreachable Code & Eliminates some unreachable code. &
 Filter/Inner Product & Yes \\
Lift Functions & Moves all functions to the top-level. & Key & Yes \\
Drop Redundant Nodes & Eliminates unnecessary nodes/nesting & Filter & No \\
Flatten Expressions & Removes nesting from expressions. & Key & Yes \\
Compress Atomic Nodes & Atomizes nested atom nodes & Amend & No \\
Propagate Constants & Inlines all references to literal values. &
 Amend/Rank & Yes \\
Fold Constants & Converts constant expressions to literals & Amend & No \\
Compress Expressions & Converts expression sub-trees into single nodes. & 
 Amend & No \\
Record Final Return Value & Records the value returned by each function. & 
 Key & No \\
Normalize Values Field & Normalizes the shape and size of the values field. &
 Amend & No \\
Lift Type-checking & Infers some type information at compile time &
 Power Limit/Rank & No \\
Allocate Value Slots & Does a form of frame allocation for variables &
 Key & No \\
Anchor Variables & Resolves lexically scoped variables &
 Key & Yes \\
Record Live Variables & Records the variables that are live at each point of 
 execution. & Key & No \\
Fuse Scalar Loops & Identifies Fusion opportunities and fuses expressions &
 Key & No \\
Type Specialization & Specializes each function for a series of potential inputs. 
 & Key & No \\
\end{tabular}
\caption{A listing of some compiler passes in the Co-dfns compiler and their 
 relationship with the Key operator and associated tree computation techniques}
\label{tab:passes}
\end{table*}

\section{Case Studies}

The following compiler passes are often found in any suitably complex 
language being compiled, and represent a difficult problem for anyone attempting 
to do data-parallel computation. Indeed, they were the primary motivating 
factor for the development of the node coordinate matrix and its associated 
techniques. They demonstrate how we can operate on highly nested structures, 
usually with the intent of eliminating that structure to more closely map 
the intended output language.

These case studies come from the Co-dfns compiler, which is compiling a 
lexically scoped language that contains both nested functions and 
expressions.

\subsection{Function Lifting}

@@

The final task to complete function lifting of each function
body before lifting is to shift the depths in the depth vectors
to correspond to those of a function lifted to the top-level, and
to attach a function node to the top of each of the bodies. At that
point, we simply recombine all of the newly created functions into
a single top level.

\subsection{Expression Flattening}

@@

The only extra issue involved here is to ensure that the order of
evaluation matches. In our case, we are assuming that the order of
evaluation is right to left, which means that the above order is
actually backwards of the desired order. During recombination, we
simply reverse these orders and this fixes that problem. More work
would be required to take into consideration a specific precedence
hierarchy.

\section{Other Passes}

Table \ref{tab:passes} contains a selection of passes from the
Co-dfns compiler and breaks down their features based on how they
traverse the tree and whether or not they use the Key operator or
Node Coordinates. They are oriented with front-end passes higher in
the table and passes further down the compiler chain at the bottom
of the table. These passes give a good idea of how the Key operator
plays a part in the complete compiler as well as how Node Coordinates
help to deal with complex sub-tree selections.

@@

In short, the goal of the structure and design of the compiler passes
given in table \ref{tab:passes} is to utilize Node Coordinates to
remove as quickly as possible the nested structure of the AST into
a flatter form that does not require complex analysis to traverse.

\section{The Co-dfns Compiler}

The Co-dfns compiler is an experimental commercial compiler funded by 
Dyalog, Ltd. It is intended to provide increased scaling and performance
for APL programmers using the dfns syntax (lexically scoped, functionally 
oriented APL). These techniques form one of the foundational elements to 
the architecture of the compiler, which is built in the style of a Nanopass 
compiler, with very small passes chained together through composition, with 
the added requirements that each compiler pass must be fully data parallel.

The compiler produces good code that performs closely with hand-written 
C code for the same programs on the GPU and CPU. It produces platform 
independent code, meaning that the compiler is able to target either the 
GPU or CPU from the same program source and produce reasonable performance 
on both.

The compiler itself is composed of three main parts: the core compiler, 
the parser, and the code generator. The code generator includes the runtime 
library of the compiler. Parsing APL code in parallel is a well-studied 
problem. The compiler currently uses parsing combinators to do its parsing, 
though we plan to implement the Two-by-Two parser given in the literature. 
The code generator itself is a simple parallel map over each node, as 
each node may be generated independent from the other nodes, leading to a 
fully parallel generator, though there are some branching elements in the 
generator that still remain, currently.

The core of the Co-dfns compiler is written entirely using function trains 
and parallel operations in the style described above. We use these techniques 
extensively throughout the compiler, and the code is publicly available 
at our repository. The core of the compiler, including some of the large 
tables comes in at around 250 lines of code. Without the tables, the core 
logic comes in at around 150 lines of code.

The entire compiler, together with its runtime libraries, parser, generator, 
and core, comes in at around 1500 lines of commented code.

\section{Future Work}

Work remains on expanding the features and capabilities of the compiler, 
to demonstrate the scalability of these techniques on more sophisticated 
optimizations. The current compiler represents a non-trivial instance of 
the successful application of these techniques, but it focuses on an untyped, 
functionally oriented language. The type inference in the current version 
of the compiler is somewhat naive. More work is required to expand the 
type inferencer to include more sophisticated type inference. Likewise, 
the compiler does not currently handle error handling such as signals or 
exceptions.

Further work remains to implement these techniques on an imperative and 
object-oriented language instead of a functional one. It would also be 
educational to implement a version for a lazy language.

The performance of these techniques has not been a problem in practice, but 
we do not have a large scale understanding of the performance characteristics 
of these techniques in real life. Further implementation and analysis is 
required to understand what sort of compiler optimizations and analyses are 
necessary to compile the Co-dfns compiler efficiently for the GPU and CPU. 
We current lack a good comparison against existing techniques for the same 
language, since the current compiler is relatively unique in that no other 
compilers to our knowledge compile the same language to the same targets 
with the same sorts of optimizations. We intend to implement other paradigms 
and other languages which do have existing compiler implementations to compare 
relative performance on multiple platforms.

While we have found this method of compiler construction to be very compact 
and to permit working with a data-parallel compiler as easily or nearly as 
easily as with a standard Nanopass compiler, we have not done extensive 
studies to determine just how comparable in programmability, ease of 
maintenance, or extensibility these techniques are with established 
compiler construction methods, such as OOP Visitors, Nanopass, or the 
type-directed functional style.

\section{Related Work}

@@

J. D. Bunda and J. A. Gerth presented a method for doing table driven
parsing of APL which suggested a parallel optimization for parsing,
but did not elucidate the algorithm [7].

\section{Conclusion}

@@

We have demonstrated the technique and the core insights behind the
data structures involved. It presents a solution to a very old and
traditional problem in a very uncommon light, by eschewing the common
practices that underlie every other significant and general solution
found in modern compilers today and replacing them with an entirely
different paradigm centred on parallelism and aggregate operations.
